COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-29 TACO-paper/coffee.tex
edits

%Computational cost is a critical limitation in scientific computing, especially
%for finite element simulations. To provide one particular example we are
%particularly concerned about, it has been well established that mesh resolution
% Consequently, our aggressive optimization of
%local assembly, which may even take up to 60% of the overall FEM's execution
%time, directly impacts the performance of large-scale scientific simulations
%running on supercomputers.

@@

%JUNK: , with synchronization handled at the level of the run-time support
%Computational cost is a critical limitation in finite element simulations. High resolution in time and space is critical to the level of error in the simulations, but computational cost rises superlinearly with resolution. The result of this is that simulations are conducted at the resolution that the user can afford (wall time, allocated supercomputing units or the direct financial cost of cloud time)

@@

%JUNK: 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to

@@

% systematic optimization strategy -> discuss the importance of domain knowledge

% A python implementation of OP2, namely PyOP2, is used in the Firedrake project as an intermediate representation for FEM computations. In Firedrake all computational kernels are generated by a compiler, (the Fenics Form Compiler, FFC), whereas in OP2 applications kernels are provided by users.

\section{Background and Motivating Examples}
\label{sec:background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.

@@

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
The absence of stencils renders the element matrix computation easily
auto-vectorizable by a vendor compiler. Nevertheless,
auto-vectorization is not efficient if data are not aligned and if the
length of the innermost loop is not a multiple of the vector length
$vl$, especially when the loops are small as in local assembly.

Data alignment is enforced in two steps. Firstly, all arrays are
allocated to addresses that are multiples of $vl$. Then, two
dimensional arrays are padded by rounding the number of columns to the
nearest multiple of $vl$. For instance, assume the original size of a
basis function array is 3$\times$3 and $vl=4$ (e.g. AVX processor,
with 256 bit long vector registers and 64-bit double-precision
floats). In this case, a padded version of the array will have size
3$\times$4. The compiler is explicitly informed about data alignment
using a suitable pragma.

Padding of all two dimensional arrays involved in the evaluation of
the element matrix also allows to safely round the loop trip count to
the nearest multiple of $vl$. This avoids the introduction of a
remainder (scalar) loop from the compiler, which would render
vectorization less efficient. These extra iterations only write to the
padded region of the element matrix, and therefore have no side
effects on the final result.

@@

%More importantly, we avoid using $vl$-1 registers per each
%outer-loop-dependent variable (i.e., all $x$ terms in
%Equation~\ref{outer-product}), with $vl$ being the vector
%length. This is a considerable gain, which allows us to slice the
%iteration space into bigger tiles, implemented directly by vector
%registers.

>>>>

%Computational cost is a critical limitation in scientific computing,
%especially for finite element simulations. To provide one particular
%example we are particularly concerned about, it has been well
%established that mesh resolution Consequently, our aggressive
%optimization of local assembly, which may even take up to 60% of the
%overall FEM's execution time, directly impacts the performance of
%large-scale scientific simulations running on supercomputers.

@@

%JUNK: , with synchronization handled at the level of the run-time
%support Computational cost is a critical limitation in finite element
%simulations. High resolution in time and space is critical to the
%level of error in the simulations, but computational cost rises
%superlinearly with resolution. The result of this is that simulations
%are conducted at the resolution that the user can afford (wall time,
%allocated supercomputing units or the direct financial cost of cloud
%time)

@@

%JUNK: 1.5$\times$ over code in which only basic transformations have
%been performed, such as generalized loop-invariant code motion,
%padding, and data alignment, and up to

@@

% systematic optimization strategy -> discuss the importance of domain
% knowledge

% A python implementation of OP2, namely PyOP2, is used in the
% Firedrake project as an intermediate representation for FEM
% computations. In Firedrake all computational kernels are generated
% by a compiler, (the Fenics Form Compiler, FFC), whereas in OP2
% applications kernels are provided by users.

\section{Background and Motivating Examples}
\label{sec:background}
%Here we discuss about the computational characteristics of
%computational science kernels. Briefly cite op2 kernels. Emphasis on
%Finite Element Assembly. Generalization and formalization of a Finite
%Element Assembly kernel using quadrature representation.

@@

% Padding and Data Alignment
\subsection{Padding and Data Alignment}
The absence of stencils renders the element matrix computation easily
auto-vectorizable by a vendor compiler. Nevertheless,
auto-vectorization is not efficient if data are not aligned and if the
length of the innermost loop is not a multiple of the vector length
$\mbox{\texttt{VL}}$, especially when the loops are small as in local assembly.

Data alignment is enforced in two steps. Firstly, all arrays are
allocated to addresses that are multiples of $\mbox{\texttt{VL}}$. Then, two
dimensional arrays are padded by rounding the number of columns to the
nearest multiple of $\mbox{\texttt{VL}}$. For instance, assume the original size of a
basis function array is 3$\times$3 and $\mbox{\texttt{VL}}=4$ (e.g. AVX processor,
with 256 bit long vector registers and 64-bit double-precision
floats). In this case, a padded version of the array will have size
3$\times$4. The compiler is explicitly informed about data alignment
using a suitable pragma.

Padding of all two dimensional arrays involved in the evaluation of
the element matrix also allows to safely round the loop trip count to
the nearest multiple of $\mbox{\texttt{VL}}$. This avoids the introduction of a
remainder (scalar) loop from the compiler, which would render
vectorization less efficient. These extra iterations only write to the
padded region of the element matrix, and therefore have no side
effects on the final result.

@@

%More importantly, we avoid using $\mbox{\texttt{VL}}$-1 registers per each
%outer-loop-dependent variable (i.e., all $x$ terms in
%Equation~\ref{outer-product}), with $\mbox{\texttt{VL}}$ being the vector
%length. This is a considerable gain, which allows us to slice the
%iteration space into bigger tiles, implemented directly by vector
%registers.
